---
title: "Using multiple R versions on Linux"
author: "Alex M. Chubaty"
date: "20 Sep 2022"
output: 
  pdf_document: 
    highlight: pygments
    toc: yes
    toc_depth: 3
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Unlike on Windows, most R users on linux tend to work with a single version of R -- whichever version is available in their software repositories.
With a little bit of work, it is possible to use multiple versions of R on Linux; however, you will need an account with `sudo` privileges to set this up.

The most common use case is to install R-devel alongside the current R version in order to facilitate package checking and testing prior to CRAN submission.
Additionally, you may wish to maintain older versions of R to maintain compatibility and reproducibility of old code/scripts, or to rerun old analyses.

# Using Docker

This is the preferred solution, as it is self-contained.
Every time you start an instance of R in a Docker container, it starts in a "factory fresh" state, which means you will need to install additional packages into each new instance or save (commit) your changes for later reuse.
The advantages here are that packages installed for each project are kept separate, ensuring proper reproducibility and portability.
(Although [`renv`](https://rstudio.github.io/renv/) is also good for maintaining separate package libraries per project, it doesn't ensure system dependency separation.)

Please note that this guide is intended only to get you started with using R with Docker, not to be a full tutorial for more advanced Docker use.
See the official [Docker documentation](https://docs.docker.com/) and the [rocker project](https://github.com/rocker-org/rocker) for more advanced usage, including using containers with Rstudio Server installed[^1] or using a virtual X server.

[^1]: https://github.com/rocker-org/rocker-versioned

## Older R versions

1. Install Docker following the instructions [here](https://docs.docker.com/engine/installation/).

2. Download the image for the version of R you wish to use:

    ```bash
    docker pull r-base:3.6.3
    ```

3. Start a container running the R version of your choice:

    ```bash
    docker run -it --rm r-base:3.6.3
    ```
    
    \* NOTE: the `--rm` flag tells docker to remove the container after use.
    If you wish to keep the container for reuse later (*e.g.*, so you don't need to reinstall packages, etc.), then omit this.

### Docker images wth Rstudio

Alternatively, there are docker images that include Rstudio server built-in, so you can run a container and connect to it through your browser instead of having to work in a terminal.

1. *E.g.*, start a versioned container running Rstudio that includes geospatial libraries:

    ```bash
    docker run --rm -ti -p 8787:8080 rocker/geospatial:4.0.3
    ```
    
2. Look for the user password to appear (in red) once it's done setting up the container.

3. Open a web browser and go to `localhost:8080` to get to the Rstudio instance running in the docker container.

4. Log in with username `rstudio` and the password from step 2.

## R-devel

1. Install Docker following the instructions [here](https://docs.docker.com/install/linux/docker-ce/ubuntu/).
2. Download the latest R-devel image:

    ```bash
    docker pull rocker/drd
    ```
3. Start a container running R-devel:

    ```bash
    docker run -it --rm rocker/drd Rdevel
    ```
    
    \* NOTE: the `--rm` flag tells docker to remove the container after use.
    If you wish to keep the container for reuse later (*e.g.*, so you don't need to reinstall packages, etc.), then omit this.

\newpage

# Metal installation

As noted above, using Docker is likely a better solution, as this approach doesn't maintain separation between packages installed for different minor versions.

## Older R versions

1. Install prerequisites for building R from source:

    ```bash
    sudo apt build-dep r-base
    ```

2. Create directories to keep the R source code:

    ```bash
    mkdir ~/R/src
    ```

3. Get the source code for the version of R you wish to install:

    R 3.y.z:
    
    ```bash
    cd ~/R/src/
    RVERSION=3.6.3
    wget https://cran.r-project.org/src/base/R-3/R-${RVERSION}.tar.gz
    tar -xvzf R-${RVERSION}.tar.gz
    ```

    R 4.y.z:
    
    ```bash
    cd ~/R/src/
    RVERSION=4.0.5
    wget https://cran.r-project.org/src/base/R-4/R-${RVERSION}.tar.gz
    tar -xvzf R-${RVERSION}.tar.gz
    ```

4. Install:

    ```bash
    cd ~/R/src/R-${RVERSION}
    ./configure --prefix=/usr/local/lib/R-${RVERSION} --enable-R-shlib --with-blas --with-lapack
    make
    sudo make install
    ```

5. Create symlink:

    ```bash
    sudo ln -s /usr/local/lib/R-${RVERSION}/bin/R /usr/local/bin/R-${RVERSION}
    ```

6. Run a specific R version:

    From the commandline, simply do (according to `${RVERSION}`):
    
    ```bash
    R-3.6.3
    # or
    R-4.0.5
    ```

## R-devel

1. Install prerequisites for building R from source:

    ```bash
    sudo apt build-dep r-base
    sudo apt install ccache subversion xorg-dev
    ```

2. Create directories to keep the R-devel source code and to `make` it:

    ```bash
    mkdir -p ~/R
    mkdir -p ~/GitHub/r-config/R-devel/src/
    ln -s ~/GitHub/r-config/R-devel/src ~/R/src
    ```

3. Get the latest version of R-devel from the subversion repository:

    ```bash
    cd ~/R/src
    svn co https://svn.r-project.org/R/trunk r-devel/R 
    ```

4. Link the recommended packages for building R:

    ```bash
    cd ~/R/src/r-devel/R
    bash ./tools/rsync-recommended
    bash ./tools/link-recommended
    ```

5. Use the installation script in `R-devel/scripts/build-R-devel.sh`:

    ```{r comment=''}
    cat(readLines("R-devel/scripts/build-R-devel.sh"), sep = '\n')
    ```

6. Give the script execute permissions:

    ```bash
    chmod a+x ~/GitHub/r-config/R-devel/scripts/build-R-devel.sh
    ```

7. Make and install R-devel:

    ```bash
    mkdir -p ~/R/src/R-devel-build
    bash ~/GitHub/r-config/R-devel/scripts/build-R-devel.sh
    
    cd ~/R/src/R-devel-build
    sudo make install
    ```

8. Create custom script to launch R-devel in `R-devel/scripts/R-devel.sh`:

    ```{r comment=''}
    cat(readLines("R-devel/scripts/R-devel.sh"), sep = '\n')
    ```

    Note that this keeps the R-devel package library separate from your regular R libraries.
    
    Be sure to give the script execute permissions:
    
    ```bash
    chmod a+x ~/GitHub/r-config/R-devel/scripts/R-devel.sh
    ```

9. Create custom script to launch Rscript-devel in `R-devel/scripts/Rscript-devel.sh`:

    ```{r comment=''}
    cat(readLines("R-devel/scripts/Rscript-devel.sh"), sep = '\n')
    ```


    Note that this keeps the R-devel package library separate from your regular R libraries.
    
    Be sure to give the script execute permissions:
    
    ```bash
    chmod a+x ~/GitHub/r-config/R-devel/scripts/Rscript-devel.sh
    ```

10. Create symlinks to launch R-devel and Rscript-devel:

    ```bash
    sudo ln -s ~/GitHub/r-config/R-devel/scripts/R-devel.sh /usr/local/bin/R-devel
    sudo ln -s ~/GitHub/r-config/R-devel/scripts/Rscript-devel.sh /usr/local/bin/Rscript-devel
    ```

11. Create a script used to update R-devel in `R-devel/scripts/update-R-devel.sh`:

    This could be turned into a `cron` job.
    
    ```{r comment=''}
    cat(readLines("R-devel/scripts/update-R-devel.sh"), sep = '\n')
    ```

    Be sure to give the script execute permissions:
    
    ```bash
    chmod a+x ~/GitHub/r-config/R-devel/scripts/update-R-devel.sh
    ```

12. Run R-devel:

    From the commandline, simply do:
    
    ```bash
    R-devel
    ```

\newpage

## Switching R versions for use with RStudio

RStudio will check for an environment variable, `RSTUDIO_WHICH_R`, which can be set to override using the installed system version of R.

### Rstudio desktop

```bash
export RSTUDIO_WHICH_R=/usr/local/bin/R-3.6.3
rstudio &
```

### Rstudio server

As `root` user:

```bash
umask 0022
echo "export RSTUDIO_WHICH_R=/usr/local/bin/R-3.6.3" > /etc/profile.d/rstudio.sh
```

# References

- <https://stackoverflow.com/a/24019938/1380598>
- <https://stat.ethz.ch/pipermail/r-sig-debian/2012-August/001937.html>
- <http://singmann.org/installing-r-devel-on-linux/>
- <https://hub.docker.com/r/rocker/drd/~/dockerfile/>
- <https://support.rstudio.com/hc/en-us/articles/218004217-Building-R-from-source>
- <https://community.rstudio.com/t/use-a-different-r-version-temporarily-in-rstudio/20848/8>
- <https://solutions.rstudio.com/sys-admin/advanced-config/environment-variables/>
